---
title: "Short note about tidyeval"
author: "Adi Sarid"
date: 2019-01-20T12:00:00-00:00
categories: ["R"]
tags: ["tidyeval"]
---

Following Jenny Bryan's talk on tidyeval in the last rstudio::conf 2019, I decided to write this short note (mainly as a reminder to myself).

# What is tidyeval?

Tidy evaluation, or non standard evaluation, allows us to pass column names between functions. This is the "classic" behaviour of most tidyverse functions. For example, we use:

```{r tidyverse simple example, warning=FALSE, message=FALSE}

library(tidyverse)

mtcars %>% 
  select(mpg, cyl)

```

The two variables were selected out of the `mtcars` data set, and we specified them as names without using any quotation marks. They are symbolic, not characters (although they could also be specified as characters, `select` is smart enough that way).

But assume we want to pass variables "tidy style" between functions which do different operations. 

# Variation one - a basic example

We'll start simple: a function which has two parameters. The first parameter is a dataset. The second parameters is a grouping variable. All other variables in the data set will have their mean computed using `summarize_all`.

```{r variation one}
test1 <- function(dataset, groupby_vars){
  grouping_vars <- enquo(groupby_vars)
  dataset %>% 
    group_by(!! grouping_vars) %>%
    summarize_all(funs(mean(.))) %>%
    return()
}

mtcars %>%
  select(cyl:carb) %>%
  test1(groupby_vars = cyl)
```

We can see that `mtcars` was grouped by `cyl` which was passed as a name (not characters). The function `test1` took it, then `enquo()`-ed it, and eventually used it in the tidy chain using `!!`. 
The function `enquo` turns the input into a "quosure". Then the `!!` "uses" the quosure to select the proper variable from mtcars.

# Passing arguments using `...`

A slightly more complex situation is passing multiple arguments to the function. Assume that this time we want to construct a function which gets one input by which to group by, and what are the variables to be summarized:

```{r using dots}
test2 <- function(dataset, groupby_vars, ...){
  grouping_vars <- enquo(groupby_vars)
  dataset %>% 
    group_by(!! grouping_vars) %>%
    summarize_at(vars(...), funs(mean(.))) %>%
    return()
}

mtcars %>%
  select(cyl:carb) %>%
  test2(groupby_vars = cyl, disp:drat)
```

What happend is that `test2` treats the grouping variable the same way that `test1` treated it, but it also passed along the variables `disp:drat`.

# Maximum flexibility - multiple `enquo()`s

Sometime passing the dots, i.e., `...` is not enough.
For example, if we want specify behaviour for different columns of the data frame (e.g., compute the mean for some and the std for others). In such cases we need a more flexible version. We can extend the flexibilty of this approach using multiple `enqou()`s.

```{r multiple enquos}
test3 <- function(dataset, groupby_vars, computemean_vars, computestd_vars){
  grouping_vars <- enquo(groupby_vars)
  mean_vars <- enquo(computemean_vars)
  std_vars <- enquo(computestd_vars)
  dataset %>% 
    group_by(!! grouping_vars) %>%
    summarize_at(vars(!!mean_vars), funs(mean(.))) %>%
    left_join(dataset %>%
                group_by(!! grouping_vars) %>%
                summarize_at(vars(!!std_vars), funs(sd(.))))
}
mtcars %>% 
  test3(groupby_vars = cyl, disp:drat, wt:carb)
```

In the resulting table, the first column `cyl` is the grouping variable, columns `disp` through `drat` have the mean of the corresponding variables, and columns `wt` through `carb` have their standard deviation computed.

# Additional uses of tidy evaluation

This evaluation is very useful when building flexible functions, but also when using the `ggplot2` syntax within functions, and more so when using Shiny applications, in which input parameters need to go in as grouping or as plotting parameters.

However, this is a topic for a different post.

# Conclusions

Tidy evaluation empowers you with great tools - it offers a great degree of flexibilty, but it's a bit tricky to master.

My suggestion is that if you're trying to master tidy evaluation, just think about your use case: which of the three variations presented in this post it resembles too? 

Work your way up - from the simplest version (if it works for you) and up to the complex (but most flexible) version.