---
title: 'Purrring progress bars (adding a progress bar to `purrr::map`)'
author: "Adi Sarid"
date: 2019-01-25T12:00:00-00:00
categories: ["R"]
tags: ["purrr", "progress bars"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Important note (2023-04-25):** In `purrr` version 1.0.0 the capability of integral progress bars was introduced. Just use the argument `.progress=TRUE` in the `map` function. See the documentation [here](https://purrr.tidyverse.org/reference/progress_bars.html). The rest of this post remains unchanged, even though it is no longer the optimal way to solve this issue.

With all the functional programming going on (i.e., `purrr::map` and the likes), there is at least one thing that I found missing: progress bars. The `plyr::do` function had a nice looking progress bar open up by default if the operation took more than 2 seconds and had at least two more to go (as per Hadley's description in [Issue#149 in tidyverse/purrr](https://github.com/tidyverse/purrr/issues/149)).

~~The issue is still open, for the time of writing these lines, and will probably be solved sometime in the near future as a feature of `purrr::map`.~~

Personally, I like @cderv's elegent solution suggested at that same [github issue](https://github.com/tidyverse/purrr/issues/149).

Here is an example implementation for reading multiple files within a directory and combining them into a single tibble while showing a progress bar when reading the files. The file reading is very similar to what was suggested in [this post](https://serialmentor.com/blog/2016/6/13/reading-and-combining-many-tidy-data-files-in-R).

```{r read multiple files with progress bars, eval=FALSE}
library(purrr)
library(readr)
library(dplyr)

# directory from which to read a bunch of files (the example here uses csv)
file_list <- dir(path = "PATH_TO_DIRECTORY", pattern = ".csv")

# define reading function which includes the progress bar updates and printing
read_with_progress <- function(filename){
  pb$tick()$print()
  data_read <- read_csv(filename)
  # you can add additional operations on data_read, or 
  # decide on entirely different task that this function should do.
}

# create the progress bar with a dplyr function. 
pb <- progress_estimated(length(file_list))
res <- file_list %>%
  map_df(~read_with_progress(.))
```

That's it. You're set to go with a cool progress bar which will print out something like this while the operation is carried out:
```
|=====================================           |80% ~23 s remaining
```
