---
title: Assessing relationships with correlograms
author: Adi Sarid
date: '2023-12-22'
slug: assessing-relationships-with-correlograms
categories:
  - R
tags:
  - Correlation
subtitle: ''
summary: ''
authors: []
lastmod: '2023-12-22T16:49:06Z'
featured: no
draft: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
editor_options: 
  markdown: 
    wrap: 72
---

We often find ourselves with a complex dataset containing numerous
variables. One of the initial steps in the discovery phase - the initial
analysis where you get familiar with the data - is using correlations to
understand the relationships between the variables.

A good tool for getting a quick glimpse of these relationships is a
**correlogram chart**, which is the hero of this post. But before that,
a few words about correlations.

## What is a correlation coefficient?

A correlation coefficient (denoted as ⍴) is a statistical measure
indicating if a statistical relationship (specifically linear) exists
between two variables (features of the data).

Correlation ranges from -1 and up to +1. The extremes indicate a strong
relationship between two variables:

-   A correlation coefficient of -1 (or close to it) indicates that when
    one variable increases, the other decreases (or vice versa)

-   A correlation coefficient close to +1 indicates that the two
    variables increase or decrease together

-   A correlation coefficient close to 0 indicates that no (linear)
    relationship was found (there might be a different relationship as
    we'll see in a moment).

## Correlation "toy example"

To illustrate, here are a few scatter plots indicating the relationship
between two variables (represented by the x-axis and the y-axis).

```{r example for correlations, warning=FALSE, message=FALSE}

library(tidyverse)
set.seed(0)
tot_samp <- 100
base_x <- runif(tot_samp)

corr_examples <- tibble(x = base_x) %>% 
  mutate(high_corr = x + rnorm(tot_samp, sd = 0.2),
         high_inve = -x + rnorm(tot_samp, sd = 0.2),
         low_corr1 = runif(tot_samp) - 0.5,
         low_corr2 = x * if_else(runif(tot_samp) < 0.5, 1, -1))

cor(corr_examples)[,1]

corr_examples %>% 
  pivot_longer(-x, values_to = "y", names_to = "correlation") %>% 
  mutate(correlation = fct_recode(correlation,
                                  "Positive relationship (⍴≈0.8)" = "high_corr",
                                  "Negative relationship (⍴≈-0.8)" = "high_inve",
                                  "No relationship (⍴≈0)" = "low_corr1",
                                  "Non-linear relationship (⍴≈0)" = "low_corr2")) %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  facet_wrap(~correlation) + 
  theme_bw() + 
  ylab("Y Variable") + 
  xlab("X Variable") + 
  ggtitle("Illustration of variable relationships and correlation coefficient")
```

These examples were engineered to show various types of relationships
and how they are reflected in the correlation coefficient. The two upper
charts illustrate strong relationships: either positive (upper-left) or
negative (upper-right).

The bottom-left chart shows no-relationships (these are just random
numbers on the x and y-axis), and the bottom-right illustrates an
interesting relationship: the two variables are not correlated (a
correlation coefficient close to 0), but they are surely related to one
another. The y-value can either be x or -x. A simple transformation of y
(absolute value), would yield a correlation coefficient of 1 instead of
0. This last example also shows the limitation of using a correlation
coefficient, having said that though, let's use a correlogram to inspect
the variables.

## Correlograms to the rescue

A correlogram is a powerful tool that allows us to get a quick glimpse
of the correlation coefficients of the data features visually. In our
toy example, it looks like this:

```{r a correlogram}
corr_examples %>% 
  set_names(c("Original variable", "Positive", "Negative", "No relationship", "Non-linear")) %>% 
  cor() %>% 
  ggcorrplot::ggcorrplot(ggtheme = theme_bw(), 
                         title = 'Correlogram of the variables in our "toy example"', 
                         colors = c("#FC8D59", "#FFFFFF", "#91CF60"))
```

Using the correlogram, we can immediately see the relationships between
our original variable, the positive and negative relationship variables.
We can see that there are two variables with low correlation.

Now, let's move to a more interesting and realistic example -
illustrating movie-genera relationships.

## Relationships of movie-genres

The IMDb has some interesting data for download (see [this
link](https://datasets.imdbws.com/)). Specifically, one of the files
they have made available is called `title.basics.tsc.gz` which contains
titles with information such as release dates, runtime (duration in
minutes), and associated genres. Since the original file is very large
in size, I have rearranged it to contain just the genres, with 0-1
coding (if a title is associated with a specific genre, the coding will
be 1, otherwise, it is 0).

```{r preprocessing imdb data, eval=FALSE}
library(tidyverse)

# Read the data
imdb_titles <- read_tsv('data.tsv')

# Have a look on what we have
glimpse(imdb_titles)

# Rework the data to extract the genres
imdb_genres  <- imdb_titles |>
   select(originalTitle, genres) |>
   separate(genres, into = c("genre1", "genre2", "genre3"), sep = ',')

# Make the data into a long format
imdb_genres_long <- imdb_genres |>
   pivot_longer(cols = -originalTitle, names_to = "gen_loc_rm", values_to = "genre") |>
   mutate(genre_assoc = 1) |>
   filter(!is.na(genre))

# Have a look on available genres in the data
imdb_genres_long  |>
   distinct(genre)

# Make it wide again - as preperation to the correlogram
imdb_genres_wide <- imdb_genres_long |>
   select(-gen_loc_rm) |>
   distinct(originalTitle, genre, .keep_all = TRUE) |>
   pivot_wider(id_cols = originalTitle, 
               names_from = genre, 
               values_from = genre_assoc, 
               values_fill = 0) |>
   select(-`\\N`)

# Export the result to preserve for use
# This one exports the entire data with title names
# write_csv(imdb_genres_wide, "imdb_genre_association.csv")
# This one exports just the 0-1 encoding without title names (takes less space)
arrow::write_parquet(imdb_genres_wide |> select(-originalTitle), 
   "imdb_genre_association_no_name.parquet")

```

```{r reading the data}
imdb_genres_coding <- arrow::read_parquet("data/imdb_genre_association_no_name.parquet")

glimpse(imdb_genres_coding)
```

The following correlogram shows the relationships between genres. It is
also "clustered" in the sense that similar genres are re-arranged to
appear as blocks.

```{r genre correlogram}
imdb_cor <- cor(imdb_genres_coding)
ggcorrplot::ggcorrplot(imdb_cor,
                       title = "Correlogram of movie genres",
                       hc.order = TRUE,
                       ggtheme = theme_bw(),
                       colors = c("#D73027", "#FFFFFF", "#1A9850"),
                       tl.cex = 10, 
                       tl.srt = 90) + 
  theme(axis.text.x = element_text(vjust = 0.5))
```

Although the correlations are not very strong between most genres (the
highest are around 0.4 between action and adventure), the chart
immediately highlights some positive relationships (partial list):

-   Action, Adventure, and Animation are related

-   History, Biography, and Documentary

-   Crime and Mystery

-   Romance and Drama

-   Talk-shows and News

-   Thriller and Horror (and also Crime and Mystery)

We can also see some negative (opposite) relationships such as:

-   Adult vs. Comedy

-   Documentary vs. Comedy

-   Drama vs. Documentary

## Conclusions

In this post, I discussed correlations in general and illustrated how
different relationships are expressed in positive, negative, or
sometimes even low correlation (when the relationship is non-linear).

We've seen how correlogram charts can be a strong visual tool for
quickly glancing at a large data set and highlighting some
relationships.

Correlograms are usually only one of the initial steps - they give us
some directions, but most of the time further research is required.
